---
title: "curatedAnVILData -- early sketch"
shorttitle: "outlier dytection methods"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{curatedAnVILData -- early sketch}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Overview

Our objective in this vignette is to illustrate the use of Bioconductor's MultiAssayExperiment
for representing and querying a large AnVIL resource: GTEx.  The steps are

- initialize an AnVIL/Terra workspace
- use gen3.theanvil.io to produce a PFB extract, export to Terra using the UI
    - in this example, extract records with ancestry "Black or African American" or "White"
- use the Bioconductor AnVIL package and R programming to explore the GTEx data

The curatedAnVILData package is currently private as it includes metadata that
are controlled-access at this time.  See the workspace description for additional
details about replicating these computations.


## Basic inspection

Load packages.

```{r begin}
suppressPackageStartupMessages({
library(AnVIL)
library(rtracklayer)
})

```

Get workspace name.

```{r ch3, eval=TRUE}
avworkspace()
```

List available tables.

```{r ch4, eval=TRUE}
ta = avtables()
ta
```

Extract the 'sequencing' table.


```{r ch2a, eval=TRUE}
seqtab = avtable(table="sequencing")
```

Its dimensions are `r nrow(seqtab)` by `r ncol(seqtab)`.

Data formats in use, and the associated type labels:


```{r ch5, eval=TRUE}
table(seqtab$`pfb:data_type`, seqtab$`pfb:data_format`)
```

A 'data category' is also recorded.

```{r ch6, eval=TRUE}
table(seqtab$`pfb:data_category`, seqtab$`pfb:data_format`)
```

## Demographics and tissue type

```{r ch7, eval=TRUE}
subj = avtable(table="subject")
table(subj$`pfb:ancestry`)
```

```{r ch9, eval=TRUE}
if (!exists("samp")) samp = avtable(table="sample") # slow
table(samp$`pfb:tissue_type`)
```

Filter to lung samples.

```{r ch11, eval=TRUE}
library(dplyr)
lungsamp = samp |> filter(`pfb:tissue_type` == "Lung")
dim(lungsamp)
```

Features available in this filtered table:

```{r ch12, eval=TRUE}
names(lungsamp)
```

What attributes are shared between sample table and sequencing table,
or sequencing table and subject table?

```{r ch13, eval=TRUE}
intersect(names(lungsamp), names(seqtab))
```
```{r ch13a}
intersect(names(seqtab), names(subj))
```

How many identifiers are common to sample, sequencing, and subject tables?

```{r ch15, eval=TRUE}
length(intersect(lungsamp$`pfb:subject`, subj$`subject_id`))
```

```{r ch16, eval=TRUE}
length(intersect(seqtab$`pfb:sample`, samp$`sample_id`))
```

We need to rename an identifier variable.

```{r ch17, eval=TRUE}
mm = mutate(lungsamp, subject_id=`pfb:subject`)
```

Join sample and subject tables.

```{r ch18, eval=TRUE}
lsasu = left_join(mm, subj, by="subject_id")
```

This leads to `r nrow(lsasu)` records with `r ncol(lsasu)` fields.

There are `r length(unique(lsasu$subject_id))` unique subject identifiers.

Modify the identifier field name in the sequencing table.

```{r ch21, eval=TRUE}
sqsq = mutate(seqtab, sample_id=`pfb:sample`)
```

Join to the sample-subject table.

```{r ch22, eval=TRUE}
lsqsasu = right_join(sqsq, lsasu, by="sample_id")
```

Now there are `r length(unique(lsqsasu$subject_id))` distinct
subjects available with sequencing data.




```{r ch25, eval=TRUE}
table(lsqsasu$`pfb:data_format`)
```

```{r ch26, eval=TRUE}
lbw = lsqsasu |> filter(`pfb:data_format`=="bigWig")
```


```{r ch27, eval=TRUE}
dim(lbw)
```


```{r ch28, eval=TRUE}
lbs_drs = lbw %>% select(`pfb:ga4gh_drs_uri`)
head(lbs_drs,3)
```

```{r ch29, eval=TRUE}
lbs_gs = drs_stat(lbs_drs[[1]]) # data.frame unwrapping
```


```{r ch30, eval=TRUE}
head(lbs_gs,3)[,1:5]
```


```{r ch31, eval=TRUE}
chk = lbs_gs[1,4][[1]]
chk
```

It would be nice to query the BigWig directly.  But the bucket is requester pays, and we may need enhancements to rtracklayer.


```{r ch32, eval=TRUE}
gsutil_cp(chk, Sys.getenv("WORKSPACE_BUCKET"))
```


```{r ch33, eval=TRUE}
gsutil_ls(Sys.getenv("WORKSPACE_BUCKET"))
```


```{r ch34, eval=TRUE}
targ = "GTEX-13X6K-1626-SM-7EWCX.Aligned.sortedByCoord.out.patched.md.bigWig"
if (!file.exists(targ)) gsutil_cp(chk, ".")
```


```{r ch35, eval=TRUE}
dir()
```


```{r ch36, eval=TRUE}
library(rtracklayer)
import(dir(patt="GTEX-")[1])
```

# Session information

```{r dosess}
sessionInfo()
```
